Cover Times of Random Walks

### 2019-04-05 paper.tex
finished first section

Let $T_{j}$ be the number of steps until the first visit to $j$.
Formally,
\begin{align}
T_{j} &= \textrm{min} \left\{t : X_t = j \right\}. \nonumber
\end{align}
Let the hitting time $\hit$ be the maximum expected time between two vertices on $G$.
Formally, 
\begin{align}
\hit &= \max_{i,j \in V(G)} \E_i[T_j]. \nonumber
\end{align}
Note that $\E_i[T_j] = \E[T_j | X_0 = i]$.
Let the cover time $\cov$ be the expected time until all vertices have been
visited by the random walk $(X_t)$.
Formally,
\begin{align}
\cov &= \E [ \max_{j \in V(G)} T_j ]. \nonumber
\end{align}

@@

\subsection{Small Example}
Always possible to calculate exactly but the number of equations grows exponentially
Special graph \cite{BH94} full cover time from vertex 1
In symmetric cases, we can calculate it
In other cases, we have to rely on bounds

@@

Our goal is to determine the cover time of a simple cycle.
However, we begin by considering the cover time of a subset
of the integers on the number line from $0$ to $N$.
We then use our analysis to build our way up to a simple cycle.

Consider a random walk on the finite linear graph.
(For the finite linear graph in \cref{fig:line}, note that $n=N+1$.)
From $X_t=i$, we move with equal probability to $i+1$
and $i-1$ until we reach either $0$ or $N$.

@@

		\draw[edge] (-\N,0) -- (\N,0);
	\end{tikzpicture}
	\caption{A finite linear graph with a vertex at each integer from $0$ to $N$.}\label{fig:line}
\end{figure}

Call $e_i$ the expected number of steps until we reach either
end of our random walk from state $i$.
Our strategy is to write and solve a recurrence relation that
uses what we know about the random walk.
If $i=0,N$, we are already at the end point and $e_i=0$.
Otherwise, we take a step.
There's a half probability our step is to the right and
a half probability it is to the left.

@@

While the final line may look unnecessarily complicated,
notice that the structure of the equality lends itself to telescoping.

To solve the telescoping equations, notice that the walk is symmetric:
we could flip the walk over the vertical axis without changing $e_i$.
This makes sense because our intuition tells us that the only 
identifying feature of $i$ is its respective distances to the endpoints.
The logical conclusion is that $e_i=e_{N-i}$.
In particular, we have already seen that $e_0=e_N=0$ and it
naturally follows that $e_1=e_{N_1}$.
We use these observations to find $e_1$.

@@

\cref{thm:line} follows from \cref{eq:line} and the observation that any
linear finite graph can be shifted so that the left end point
is at $0$ and the right end point is at $N$.

\begin{theorem}
The expected number of steps from a state to either end of a finite linear graph
is the product of the distances from that state to each end point.
\label{thm:line}
\end{theorem}

We now return to cover times.
Consider a random walk on a simple cycle as in \cref{fig:cycle}.
(Note that the starting vertex is arbitrary since the graph is symmetric.)
Think of $(X_t)$ as the walk on the number line (between $-n$ and $n$)
where $X_0 = 0$.
The cover time $\cov$ of the simple cycle is the expected number of steps 
until we visit $n$ distinct integers.

@@

At time $T_{k}$, we are either at integer $a$ or $b$.
Then $\E[T_{k+1}-T_{k}]$ is the expected time until we reach $a-1$ or $b+1$.
Notice that we can shift our end points to 0 and $b+1-(a-1) = b-a + 2 = k+1$.
The expected number of steps to either $a-1$ or $b+1$
is $1\dot (k+1-1) = k$ by \cref{thm:line}.

@@

\subsection{Finite Linear Graph}
We have already seen a finite linear graph.
(For the finite linear graph in \cref{fig:line}, note that $n=N+1$).
\cref{thm:line} tells us the expected time from an arbitrary point to
one of the end points but we do not yet know the cover time.

Observe that the cover time is the earliest time that both end points 
have been visited.
We can use our earlier result to find the time to one of the end points
from an arbitrary position but we still need to find the expected time
from that end point to the other.

Consider $\E_i[T_N]$ the expected time to $N$ from vertex $i$.
We want to find $\E_0[T_N]$. By symmetry, $\E_0[T_N] = \E_N[T_0]$.

@@

\cite{AF14}
\cite{BH94}
\cite{Bo98}
\cite{DS84}
\cite{Du11}
\cite{Gr10}
\cite{LP17}

>>>>

Let $T_{j}$ be the number of steps until the first visit to $j$.
Formally,
\begin{align}
T_{j} &= \textrm{min} \left\{t : X_t = j \right\}. \nonumber
\end{align}
Let the hitting time $\hit$ be the maximum expected time between two vertices on $G$.
Formally, 
\begin{align}
\hit &= \max_{i,j \in V(G)} \E_i[T_j]. \nonumber
\end{align}
Note that $\E_i[T_j] = \E[T_j | X_0 = i]$.
Let the cover time $\cov$ be the until all vertices have been
visited by the random walk $(X_t)$.
We typically refer to the expected cover time $\cov$.
Formally,
\begin{align}
\cov &= \E [ \max_{j \in V(G)} T_j ]. \nonumber
\end{align}

@@

\subsection{Small Example}
In general, it is always possible to find the exact cover time
of a graph using first step analysis.
We demonstrate the strategy on a small graph suggested by \cite{BH94}.

\cref{fig:example} shows our small graph and the first step analysis
for the random walk $(X_t)$ started from the top left vertex.
The black vertex indicates the current vertex of the random walk
and the empty vertices indicate the visited nodes.

For a graph with $n$ vertices, there are $n$ possible choices for the current
vertex and $2^{n-1}$ possible combinations of visited and unvisited vertices
for each one.
Of course, some states are inaccessible.
But even on \cref{fig:example} where we exploit symmetry,
there are almost a prohibitive number of cases.
It follows from our combinatorial analysis that the number of states
grows exponentially with the number of vertices.

\begin{figure}[ht]
	\centering
		\begin{tikzpicture} [
            > = stealth, % arrow head style
            shorten > = 1pt, % don't touch arrow head to node
            %auto,
            node distance = 3cm, % distance between nodes
            semithick % line style
        ]

			\tikzstyle{edge}=[draw,line width = .5pt,-,black!100]

			\node (m) at (-1.2,1.75) {
				\begin{tikzpicture}[scale = 3]
				    \tikzstyle{edge}=[draw,line width = .5pt,-,black!100]
				    \tikzstyle{current}=[draw,circle,fill=black,scale=.5]
				    \tikzstyle{visited}=[draw,circle,fill=white,scale=.5]
				    \draw[edge] (0,0) -- (0,1);
				    \draw[edge] (0,0) -- (1,0);
				    \draw[edge] (0,1) -- (1,0);
				    \draw[edge] (0,1) -- (1,1);
				    \draw[edge] (1,1) -- (1,0);
				    \node (l) at (.5, 1.1) {$\cov$};
				    \node (bl) at (0,0) {};
				    \node[current] (tl) at (0,1) {};
				    \node (br) at (1,0) {};
				    \node (tr) at (1,1) {};;
				\end{tikzpicture}
			};
			\node (a) at (2,3) {
				\begin{tikzpicture}[scale = 1]
				    \tikzstyle{edge}=[draw,line width = .5pt,-,black!100]
				    \tikzstyle{current}=[draw,circle,fill=black,scale=.5]
				    \tikzstyle{visited}=[draw,circle,fill=white,scale=.5]
				    \draw[edge] (0,0) -- (0,1);
				    \draw[edge] (0,0) -- (1,0);
				    \draw[edge] (0,1) -- (1,0);
				    \draw[edge] (0,1) -- (1,1);
				    \draw[edge] (1,1) -- (1,0);
				    \node (l) at (.5, 1.3) {a};
				    \node (bl) at (0,0) {};
				    \node[visited] (tl) at (0,1) {};
				    \node[current] (br) at (1,0) {};
				    \node (tr) at (1,1) {};;
				\end{tikzpicture}
			};
			\node (b) at (2,0) {
				\begin{tikzpicture}[scale = 1]
				    \tikzstyle{edge}=[draw,line width = .5pt,-,black!100]
				    \tikzstyle{current}=[draw,circle,fill=black,scale=.5]
				    \tikzstyle{visited}=[draw,circle,fill=white,scale=.5]
				    \draw[edge] (0,0) -- (0,1);
				    \draw[edge] (0,0) -- (1,0);
				    \draw[edge] (0,1) -- (1,0);
				    \draw[edge] (0,1) -- (1,1);
				    \draw[edge] (1,1) -- (1,0);
				    \node (l) at (.5, -.3) {b};
				    \node (bl) at (0,0) {};
				    \node[visited] (tl) at (0,1) {};
				    \node (br) at (1,0) {};
				    \node[current] (tr) at (1,1) {};;
				\end{tikzpicture}
			};
			\node (c) at (4,3) {
				\begin{tikzpicture}[scale = 1]
				    \tikzstyle{edge}=[draw,line width = .5pt,-,black!100]
				    \tikzstyle{current}=[draw,circle,fill=black,scale=.5]
				    \tikzstyle{visited}=[draw,circle,fill=white,scale=.5]
				    \draw[edge] (0,0) -- (0,1);
				    \draw[edge] (0,0) -- (1,0);
				    \draw[edge] (0,1) -- (1,0);
				    \draw[edge] (0,1) -- (1,1);
				    \draw[edge] (1,1) -- (1,0);
				    \node (l) at (.5, 1.3) {c};
				    \node (bl) at (0,0) {};
				    \node[visited] (tl) at (0,1) {};
				    \node[visited] (br) at (1,0) {};
				    \node[current] (tr) at (1,1) {};;
				\end{tikzpicture}
			};
			\node (d) at (4,0) {
				\begin{tikzpicture}[scale = 1]
				    \tikzstyle{edge}=[draw,line width = .5pt,-,black!100]
				    \tikzstyle{current}=[draw,circle,fill=black,scale=.5]
				    \tikzstyle{visited}=[draw,circle,fill=white,scale=.5]
				    \draw[edge] (0,0) -- (0,1);
				    \draw[edge] (0,0) -- (1,0);
				    \draw[edge] (0,1) -- (1,0);
				    \draw[edge] (0,1) -- (1,1);
				    \draw[edge] (1,1) -- (1,0);
				    \node (l) at (.5, -.3) {d};
				    \node (bl) at (0,0) {};
				    \node[current] (tl) at (0,1) {};
				    \node (br) at (1,0) {};
				    \node[visited] (tr) at (1,1) {};;
				\end{tikzpicture}
			};
			\node (e) at (6,3) {
				\begin{tikzpicture}[scale = 1]
				    \tikzstyle{edge}=[draw,line width = .5pt,-,black!100]
				    \tikzstyle{current}=[draw,circle,fill=black,scale=.5]
				    \tikzstyle{visited}=[draw,circle,fill=white,scale=.5]
				    \draw[edge] (0,0) -- (0,1);
				    \draw[edge] (0,0) -- (1,0);
				    \draw[edge] (0,1) -- (1,0);
				    \draw[edge] (0,1) -- (1,1);
				    \draw[edge] (1,1) -- (1,0);
				    \node (l) at (.5, 1.3) {e};
				    \node (bl) at (0,0) {};
				    \node[current] (tl) at (0,1) {};
				    \node[visited] (br) at (1,0) {};
				    \node[visited] (tr) at (1,1) {};;
				\end{tikzpicture}
			};		
			\node (f) at (6, 0) {
				\begin{tikzpicture}[scale = 1]
				    \tikzstyle{edge}=[draw,line width = .5pt,-,black!100]
				    \tikzstyle{current}=[draw,circle,fill=black,scale=.5]
				    \tikzstyle{visited}=[draw,circle,fill=white,scale=.5]
				    \draw[edge] (0,0) -- (0,1);
				    \draw[edge] (0,0) -- (1,0);
				    \draw[edge] (0,1) -- (1,0);
				    \draw[edge] (0,1) -- (1,1);
				    \draw[edge] (1,1) -- (1,0);
				    \node (l) at (.5, -.3) {f};
				    \node[current] (bl) at (0,0) {};
				    \node[visited] (tl) at (0,1) {};
				    \node (br) at (1,0) {};
				    \node[visited] (tr) at (1,1) {};;
				\end{tikzpicture}
			};
			\node (done) at (8, 3) {
				\begin{tikzpicture}[scale = 1]
				    \tikzstyle{edge}=[draw,line width = .5pt,-,black!100]
				    \tikzstyle{current}=[draw,circle,fill=black,scale=.5]
				    \tikzstyle{visited}=[draw,circle,fill=white,scale=.5]
				    \draw[edge] (0,0) -- (0,1);
				    \draw[edge] (0,0) -- (1,0);
				    \draw[edge] (0,1) -- (1,0);
				    \draw[edge] (0,1) -- (1,1);
				    \draw[edge] (1,1) -- (1,0);
				    \node (l) at (.5, 1.3) {0};
				    \node[visited] (bl) at (0,0) {};
				    \node[visited] (tl) at (0,1) {};
				    \node[visited] (br) at (1,0) {};
				    \node[visited] (tr) at (1,1) {};;
				\end{tikzpicture}
			};
			\node (g) at (8, 0) {
				\begin{tikzpicture}[scale = 1]
				    \tikzstyle{edge}=[draw,line width = .5pt,-,black!100]
				    \tikzstyle{current}=[draw,circle,fill=black,scale=.5]
				    \tikzstyle{visited}=[draw,circle,fill=white,scale=.5]
				    \draw[edge] (0,0) -- (0,1);
				    \draw[edge] (0,0) -- (1,0);
				    \draw[edge] (0,1) -- (1,0);
				    \draw[edge] (0,1) -- (1,1);
				    \draw[edge] (1,1) -- (1,0);
				    \node (l) at (.5, -.3) {g};
				    \node[visited] (bl) at (0,0) {};
				    \node[current] (tl) at (0,1) {};
				    \node (br) at (1,0) {};
				    \node[visited] (tr) at (1,1) {};;
				\end{tikzpicture}
			};
			\begin{comment}
		    \path[->] (m) edge node[above] {$\frac{1}{3}$} (a);	
		    \path[->] (m) edge node[below] {$\frac{2}{3}$} (b); 
		    \path[->] (a) edge[loop above] node[below] {$\frac{1}{3}$} (a);
		    \path[->] (a) edge node[above] {$\frac{2}{3}$} (c);
		    \path[->] (b) edge[bend left] node[below] {$\frac{1}{2}$} (d);
		    \path[->] (b) edge node[below] {$\frac{1}{2}$} (e);
		    \path[->] (c) edge[bend left] node[above] {$1$} (e);
		    \path[->] (d) edge node[below] {$\frac{1}{3}$} (e);
		    \path[->] (d) edge[bend left] node[below] {$\frac{1}{3}$} (b); 
		    \path[->] (d) edge node[below] {$\frac{1}{3}$} (f);
		    \path[->] (e) edge[bend left] node[above] {$\frac{1}{3}$} (c);
		    \path[->] (e) edge[loop above] node[below] {$\frac{1}{3}$} (e);
		    \path[->] (f) edge[bend left] node[below] {$\frac{1}{2}$} (g);
		    \path[->] (g) edge[bend left] node[below] {$\frac{2}{3}$} (f);
			\path[->] (e) edge node[above] {$\frac{1}{3}$} (done);
			\path[->] (f) edge node[below] {$\frac{1}{2}$} (done);
			\path[->] (g) edge node[right] {$\frac{1}{3}$} (done);
			\end{comment}
			\path[->] (m) edge (a);
			\path[->] (m) edge (b);
			\path[->] (a) edge[loop above] (a);
			\path[->] (a) edge (c);
			\path[->] (b) edge[bend left] (d);
			\path[->] (b) edge (e);
			\path[->] (c) edge[bend left] (e);
			\path[->] (d) edge (e);
			\path[->] (d) edge[bend left] (b);
			\path[->] (d) edge (f);
			\path[->] (e) edge[bend left] (c);
			\path[->] (e) edge[loop above] (e);
			\path[->] (f) edge[bend left] (g);
			\path[->] (g) edge[bend left] (f);
			\path[->] (e) edge (done);
			\path[->] (f) edge (done);
			\path[->] (g) edge (done);
	\end{tikzpicture}
	\caption{First step analysis applied to an asymmetric graph with $n=4$.}\label{fig:example}
\end{figure}

If we let the label of each graph in \cref{fig:example}
denote the expected cover time from that graph, we have the following system
of equations:
\begin{align}
\cov &= 1 + \frac{1}{3}\textrm{a} + \frac{2}{3}\textrm{b} \nonumber \\
\textrm{a} &= 1 + \frac{1}{3}\textrm{a} + \frac{2}{3}\textrm{c} \nonumber \\
\textrm{b} &= 1 + \frac{1}{2}\textrm{b} + \frac{1}{2}\textrm{e} \nonumber \\
\textrm{c} &= 1 + 1\textrm{e} \nonumber\\
\textrm{d} &= 1 + \frac{1}{3}\textrm{a} + \frac{1}{3}\textrm{e} + \frac{1}{3}\textrm{f} \nonumber \\
\textrm{e} &= 1 + \frac{1}{3}\textrm{c} + \frac{1}{3}\textrm{e} + \frac{1}{3}\dot 0\nonumber \\
\textrm{f} &= 1 + \frac{1}{2}\textrm{g} + \frac{1}{2}\dot 0\nonumber \\
\textrm{g} &= 1 + \frac{2}{3}\textrm{f} + \frac{1}{3}\dot 0\nonumber \\
\end{align}

With a little linear algebra, we find that $\cov = \frac{43}{6}$.

For large graphs, the first step analysis as an approach to finding the expected
cover time is obviously intractable.
The rest of this section deals with particularly symmetric graphs
for which it is possible find a solution for an arbitrary $n$.
However, we must rely on expected cover time bounds for the vast majority of cases.

@@

\subsection{Line Graph}
A line graph is a set of vertices connected in a line.
(\cref{fig:line} is an example with $n=N+1$.)

Our goal is to determine the cover time of a line graph.
Let $(X_t)$ be the random walk started at vertex $i$.
Notice that a random walk will have covered all vertices
exactly when it has visited both endpoints.
We can use this observation to break up the cover time into the time it takes
to reach one of the endpoints and the time it takes to reach 
the other endpoint.

We will begin by considering the time $(X_t)$ takes to
visit either one of the endpoints from vertex $i$.

From $X_t=i$, we move with equal probability to $i+1$
and $i-1$ until we reach either $0$ or $N$.

@@

		\draw[edge] (-\N,0) -- (\N,0);
	\end{tikzpicture}
	\caption{A   line graph with a vertex at each integer from $0$ to $N$.}\label{fig:line}
\end{figure}

Call $e_i$ the expected number of steps until we reach either
end of our random walk from state $i$.
Our strategy is to write and solve a recurrence relation that
uses what we know about the random walk.
If $i=0,N$, we are already at the endpoint and $e_i=0$.
Otherwise, we take a step.
There's a half probability our step is to the right and
a half probability it is to the left.

@@

While the final line may look more complicated than what we started with,
notice that the structure of the equality lends itself to telescoping.

To solve the telescoping equations, observe that the walk is symmetric:
we could flip the walk over the vertical axis without changing $e_i$.
This makes sense because our intuition tells us that the only 
identifying feature of $i$ is its respective distances to the endpoints.
The logical conclusion is that $e_i=e_{N-i}$.
In particular, we have already seen that $e_0=e_N=0$ and it
naturally follows that $e_1=e_{N_1}$.
We use these observations to find $e_1$.

@@

\cref{thm:line} follows from \cref{eq:line} and the observation that any
line graph can be shifted so that the left endpoint
is at $0$ and the right endpoint is at $N$.

\begin{theorem}
The expected number of steps from a state to either end of a line graph
is the product of the distances from that state to each endpoint.
\label{thm:line}
\end{theorem}

We have the expected time until we reach one endpoint of our line graph.
Now we want to find the expected time from one endpoint to the other $\E_0[T_N]$.
By symmetry, $\E_0[T_N] = \E_N[T_0]$.

We write
\begin{align}
\E_0[T_N] &= \E_0[T_1] + \E_1[T_2] + ... + \E_{i}[T_{i+1}] + ... + \E_{N-1}[T_N]. \nonumber
\end{align}

where $\E_{i}[T_{i+1}]$ is the expected time from state $i$ to $i+1$.
Since there is only one edge incident to 0, we always move from 0 to 1.
From state $i\neq0$, we take one step in each direction with equal probability.
In half the cases, we have arrived at $i+1$.
In the other half of cases, we are in $i-1$.
It will take us $\E_{i-1}[T_i]$ steps back to $i$ and then another
$\E_i[T_{i+1}]$ steps to $i+1$.
Formally,
\begin{align}
\E_{i}[T_{i+1}] &= 1 + \frac{1}{2}(\E_{i-1}[T_i]+\E_i[T_{i+1}]) \nonumber \\
 &= 2 + \E_{i-1}[T_{i}]. \nonumber
\end{align}

We can use $\E_0[T_1] = 1$ to solve for $E_i[T_{i+1}]$:
\begin{align}
E_i[T_{i+1}] &= E_{i-1}[T_{i}] + 2 = E_{i-2}[T_{i-1}] + 4 \nonumber\\
 &= E_0[T_1] + 2(i-1) = 2i - 1. \nonumber
\end{align}

Then
\begin{align}
\E_0[T_N] &= 1 + 3 + ... + 2i - 1 + ... + 2(N-1) -1 \nonumber \\
&= N^2 \nonumber
\end{align}

Then, by \cref{thm:line}, for the random walk from state $i$ on the line graph,
\begin{align}
\cov = i(N-i) + N^2. \nonumber
\end{align}

Consider a random walk on a simple cycle as in \cref{fig:cycle}.
(Note that the starting vertex is arbitrary since the graph is symmetric.)
Let $(X_t)$ be the walk on the number line (between $-n$ and $n$)
where $X_0 = 0$.
The cover time $\cov$ of the simple cycle is the expected number of steps 
until we visit $n$ distinct integers.

@@

At time $T_{k}$, we are either at integer $a$ or $b$.
Then $\E[T_{k+1}-T_{k}]$ is the expected time until we reach $a-1$ or $b+1$.
Notice that we can shift our endpoints to 0 and $b+1-(a-1) = b-a + 2 = k+1$.
The expected number of steps to either $a-1$ or $b+1$
is $1\dot (k+1-1) = k$ by \cref{thm:line}.

@@

@@

\cite{AF14}

\cite{Bo98}
\cite{DS84}
\cite{Du11}
\cite{Gr10}
\cite{LP17}
